The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-05 paper/ms.tex
Move figures around in paper

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/cartoon.pdf}
\end{center}
\caption{Cartoon approximation showing the effect an error in the
determination of the coordinate midplane can have on orbit integration and
action estimation. The $x$-axis shows the orbital phase and the $y$-axis the
vertical height. The top gray curve depicts an example ``true'' orbit
oscillating about the true midplane (horizontal solid gray line). Consider an
observer who erroneously assumes the midplane is located at the horizontal
dashed line. Suppose this observer measures the phase-space position of the
star at two different orbital phases (teal and orange points). If the
observer were then to integrate the star's orbit using a model potential with
the erroneous midplane, they would obtain the teal and orange curves for the
star's orbit, respectively. The actions estimated from these two  erroneous
orbits would subsequently differ, both from each other and from the actions
estimated for the true orbit (in the potential with the correct midplane).
Hence an incorrect midplane in the potential model assumed will induce
phase-dependence in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

\subsection{Epicyclic Approximation} \label{ssec:epi_action}
Before turning to numerical methods, we first write down analytically the
systematic error in the actions induced from either a midplane offset or a
Galactocentric radius offset. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential (i.e. ignoring the influence of the Galactic bar and
spiral arms). We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle \citep[e.g.][]{2017MNRAS.465...76M}.

@@

\subsection{Quantification of the Midplane Effect} \label{ssec:quant}
We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the \emph{observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the observed actions induced by an
error in the Galactocentric coordinate system. We consider here the thick-disk
orbit, which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$
value (upper panels) or the $x$ value (lower panels) of each
position in the orbit, corresponding to an erroneous observer assuming a
midplane (upper) or solar radius (lower) that is off by
$100\,\pc$. We then allow an immortal observer to observe the orbit over
$1\,\Gyr$ and perform the same orbit integration procedure at each timestep,
and report the values of the actions. The computation of $J_{\phi}$ is
pristine to errors in $z$, with only numerical artifacts remaining. Only small
errors are induced in $J_R$, with the middle $90\%$ of values over the $\Gyr$
being within $\sim8\%$ of the true $J_R$. As expected, large errors are
induced in $J_z$ with a $100\,\pc$ offset in $z$, with the middle $90\%$ of
values being within $\sim43\%$ of the true $J_z$. The $x$ offset induces
uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of $\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The upper panels
consider an offset in $z$ and the lower panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equation~\eqref{eq:Ji_err_mosttime}). In the epicyclic approximation, a $z$
offset only induces an error in $J_z$ --- for all three orbits the epicyclic
approximation is a good description of the $J_z$ error. An $x$ offset induces
an error in $J_R$ and $J_{\phi}$. The error in $J_R$ is somewhat
well-described for the thin-disk and thick-disk orbits, and a poor description
for the halo orbit. For $J_{\phi}$, the epicyclic approximation is not a good
description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

@@

\begin{deluxetable*}{cccccc}
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
Milky Way\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

@@

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} (left,
center, and right panels) as a function of azimuthal angle. All calculations are done at
cosmological redshift $\z =0$. The local midplane is determined at a position
$\phi$ by taking the median height of all stars within $R=0.5\,\kpc$ and
$z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma $
error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The upper
panels show the midplane as a function of azimuth before the first encounter
near the Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur which bring the
Sagittarius-like object closer to the Solar circle, we can see that stronger
midplane variations result. In fact, one can see that an $m=2$ mode is excited
in the lower left panel at $t=6.0\,\Gyr$ --- this is consistent with Figure~17 of
\citet{2018MNRAS.481..286L}, which shows that an $m=2$ mode is excited (note
that $m=0$ and $m=1$ modes are stronger, but these are removed in our
sine-curve subtraction). The fact that the $t=6.9\,\Gyr$ panel looks
qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that the midplane variations are
produced in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/orbits.pdf}
\end{center}
\caption{The three orbits presented in Table~\ref{tab:orbits} and considered
throughout the work. We plot the thin-disk, thick-disk, and halo orbits in the left, center, and right columns, respectively. The upper
row shows a plot of $x$ vs. $y$ while the lower row shows $R$ vs. $z$.}
\label{fig:plot_orbits}
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The LSR as a function of azimuth at the
fiducial Solar circle ($R_0 = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}). Variations in $v_z$ are on the order of
$\sim5\textup{--}10\,\kms$.}
\label{fig:lsr_variations}
\end{figure*}

>>>>

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/cartoon.pdf}
\end{center}
\caption{Cartoon approximation showing the effect an error in the
determination of the coordinate midplane can have on orbit integration and
action estimation. The $x$-axis shows the orbital phase and the $y$-axis the
vertical height. The top gray curve depicts an example ``true'' orbit
oscillating about the true midplane (horizontal solid gray line). Consider an
observer who erroneously assumes the midplane is located at the horizontal
dashed line. Suppose this observer measures the phase-space position of the
star at two different orbital phases (teal and orange points). If the
observer were then to integrate the star's orbit using a model potential with
the erroneous midplane, they would obtain the teal and orange curves for the
star's orbit, respectively. The actions estimated from these two  erroneous
orbits would subsequently differ, both from each other and from the actions
estimated for the true orbit (in the potential with the correct midplane).
Hence an incorrect midplane in the potential model assumed will induce
phase-dependence in the actions estimated for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

@@

\subsection{Epicyclic Approximation} \label{ssec:epi_action}

Before turning to numerical methods, we first write down analytically the
systematic error in the actions induced from either a midplane offset or a
Galactocentric radius offset. We use the epicyclic approximation, which
assumes that the motion in the $z$ and $R$ components of the orbit are
decoupled and follow simple harmonic motion about a circular and planar
``guiding orbit'' \citep[][Section~3.2 and references
therein]{2008gady.book.....B}. We refer to the radius of this orbit (the
``guiding radius'') as $R_g$. This approximation is an excellent description
of thin disk-like orbits and a good description of thick disk-like orbits in
an axisymmetric potential (i.e. ignoring the influence of the Galactic bar and
spiral arms). We also make the assumption of a perfectly flat circular
velocity curve with velocity $v_c$, a good approximation near the Solar circle \citep[e.g.][]{2017MNRAS.465...76M}.

@@

\subsection{Quantification of the Midplane Effect} \label{ssec:quant}

\begin{figure*}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_one_orbit.pdf}
\end{center}
\caption{The artificial phase-dependence in the observed actions induced by an
error in the Galactocentric coordinate system. We consider here the thick-disk
orbit, which has actions of $(J_R, J_{\phi}, J_z) = (37.9, -1520,
7.0)\,\actunit$ and $z_{\text{max}}=850\,\pc$ (see Table~\ref{tab:orbits}). We
integrate the orbit according to the procedure laid out in
Section~\ref{ssec:action_comp}, and which we plot in Appendix~\ref{app:orbits}. Then, we subtract $100\,\pc$ from the $z$
value (upper panels) or the $x$ value (lower panels) of each
position in the orbit, corresponding to an erroneous observer assuming a
midplane (upper) or solar radius (lower) that is off by
$100\,\pc$. We then allow an immortal observer to observe the orbit over
$1\,\Gyr$ and perform the same orbit integration procedure at each timestep,
and report the values of the actions. The computation of $J_{\phi}$ is
pristine to errors in $z$, with only numerical artifacts remaining. Only small
errors are induced in $J_R$, with the middle $90\%$ of values over the $\Gyr$
being within $\sim8\%$ of the true $J_R$. As expected, large errors are
induced in $J_z$ with a $100\,\pc$ offset in $z$, with the middle $90\%$ of
values being within $\sim43\%$ of the true $J_z$. The $x$ offset induces
uncertainties in $J_R$, $J_{\phi}$, and $J_z$ of $\sim21\%$, $\sim3\%$, and $\sim3\%$.}
\label{fig:one_orbit_wrong_ref}
\end{figure*}

We quantify how a systematic error in the Galactocentric coordinate system
induces phase-dependence in the \emph{observed} actions. We will consider three
orbits in the model potential described in Section~\ref{ssec:action_comp} that
are typical of stars in the stellar thin disk, stellar thick disk, and the
stellar halo. We summarize their initial positions in phase space and the
actions computed by integrating their orbits in the correct potential in
Table~\ref{tab:orbits}. Each orbit, integrated without systematic coordinate
errors, is plotted in Appendix~\ref{app:orbits}. We will refer to these orbits
by their names (thin-disk, thick-disk, halo) henceforth.

@@

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95\uth{} minus 5\uth{} percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for the
thin, thick, and halo orbits (Table~\ref{tab:orbits}). See discussion in the
text for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the
result for $J_R$, $J_{\phi}$, and $J_z$, respectively. The upper panels
consider an offset in $z$ and the lower panels consider an offset in $x$
(equivalently, an offset in the Solar radius). In some panels we also plot as
dashed lines the epicyclic prediction of the induced action error
(Equation~\eqref{eq:Ji_err_mosttime}). In the epicyclic approximation, a $z$
offset only induces an error in $J_z$ --- for all three orbits the epicyclic
approximation is a good description of the $J_z$ error. An $x$ offset induces
an error in $J_R$ and $J_{\phi}$. The error in $J_R$ is somewhat
well-described for the thin-disk and thick-disk orbits, and a poor description
for the halo orbit. For $J_{\phi}$, the epicyclic approximation is not a good
description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

@@

\begin{deluxetable*}{cccccc}[htb!]
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work. For comparison, we also give the softening lengths for the simulated galaxies.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
Milky Way\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

@@

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} (left,
center, and right panels) as a function of azimuthal angle. All calculations are done at
cosmological redshift $\z =0$. The local midplane is determined at a position
$\phi$ by taking the median height of all stars within $R=0.5\,\kpc$ and
$z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma $
error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The upper
panels show the midplane as a function of azimuth before the first encounter
near the Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur which bring the
Sagittarius-like object closer to the Solar circle, we can see that stronger
midplane variations result. In fact, one can see that an $m=2$ mode is excited
in the lower left panel at $t=6.0\,\Gyr$ --- this is consistent with Figure~17 of
\citet{2018MNRAS.481..286L}, which shows that an $m=2$ mode is excited (note
that $m=0$ and $m=1$ modes are stronger, but these are removed in our
sine-curve subtraction). The fact that the $t=6.9\,\Gyr$ panel looks
qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that the midplane variations are
produced in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/orbits.pdf}
\end{center}
\caption{The three orbits presented in Table~\ref{tab:orbits} and considered
throughout the work. We plot the thin-disk, thick-disk, and halo orbits in the left, center, and right columns, respectively. The upper
row shows a plot of $x$ vs. $y$ while the lower row shows $R$ vs. $z$.}
\label{fig:plot_orbits}
\end{figure*}

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The LSR as a function of azimuth at the
fiducial Solar circle ($R_0 = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}). Variations in $v_z$ are on the order of
$\sim5\textup{--}10\,\kms$.}
\label{fig:lsr_variations}
\end{figure*}
